Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-05-05 WSSSPE2_report/WSSSPE2_review.tex
Minor changes.

The Second Workshop on Sustainable Software for Science: Practice and
Experiences
(WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was
held on  16 November 2014 in conjunction with the International
Conference for High Performance Computing, Networking, Storage and Analysis
(SC14)\footnote{\url{http://sc14.supercomputing.org}}. WSSSPE2 followed the
model of a general initial workshop,
WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
which co-occurred with SC13, and a focused workshop,
WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}},
which was organized in July 2014 jointly with the SciPy 
conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}}.

@@

The review process was fairly standard, where reviewers bid for papers, then an
automated system matched bids to determine assignments, and reviewers then
completed their assigned reviews (with an average of 4.9 reviews per paper, and
4.1 reviews per reviewer). This process was done through
EasyChair\footnote{http://easychair.org/}, which allowed reviewers to provide
scores on relevance and comments to the organizers, who used the information to
decide which papers to associate with the workshop, and provided the comments to
the authors to help them improve their papers.

@@

\subsection{Marlon Pierce: Patching It Up, Pulling It Forward~\cite{Pierce_poster}}
Pierce discussed how open open source is. Open software needs a diverse, openly
governed community behind it, just as it needs open licensing and a public code
repository. To probe the level of governance within open source projects, he and
his co-authors (Suresh Marru and Chris Mattmann) suggested a contest to encourage
individual developers to submit patches and pull request to projects that are
important to them. This simple mechanism will expose several governance
mechanisms, such as how easy it is for independent developers to communicate
with project leadership, how projects accept and properly license third party
contributions, and how projects make decisions such as granting source tree
write access.

@@

Of the short actionable papers that would lead to improvements for sustainable
software science, 11 submissions were categorized in the  Software Development Experiences
group. Because of the large number, we split these into two subgroups
prior to the event. Some common themes helped in this division, for
example several papers that addressed education and training issues,
including best practices and case studies, were grouped
together. Others discussed experiences with registries, developer
collectives and specific examples of successful, sustainable software
(including a valuable industry perspective).

@@

We asked ourselves, ``Where are there examples of things that are changing
because of this and how can we build momentum?'' One example demonstrates
change over time. In 2007, nanoHUB listed the academic reward structure as
a problem in an EDUCAUSE report where the authors note, ``In the future,
nanoHUB researchers are hoping to change the research culture. While they
recognize that young faculty members are unlikely to get tenure based on
their nanoHUB contributions, they hope to encourage faculty to think
beyond their own research needs to consider publishing tutorials and other
Web content in their field on the nanoHUB site.'' Fast forward seven years
later where, in a 2014 iSGTW article, quote nanoHUB PI Gerhard Klimeck, ``A
former student of mine published eight tools on nanoHUB, serving over
6,000 people with his tools. He then joined a university as a professor
and introduced nanoHUB. Use of the gateway from that university
skyrocketed; he used nanoHUB in existing classes, created new classes, and
infused it in his research. Ultimately, the professor's department head
associates his two-year rise to tenure with the notoriety and innovation
he gained through nanoHUB.''

@@

We then moved beyond our training discussion to address funding that
encourages sustainable software -- funding of both people and of projects
that create a true system to support sustainability. We called this
``institutionalized serendipity.'' Because science is increasingly reliant
on software, one participant observed that ``software development can have
much broader impact than publishing individual research, but it is not
viewed that way.'' Because of this centrality, one participant mentioned
that training ought to be called Science Carpentry rather than Software
Carpentry. We felt we were beginning to see changes in the research
community as a result of the NSF's data management plan requirements. PIs
are thinking more about data and some university libraries are offering
data repositories. We wondered if a software management plan might be
effective. ``How might funding programs need to adapt to reward good
software development practices?'' we asked ourselves. We thought about
best practices such as version control, test harnesses, mailing lists, bug
tracking, community contribution, and reuse where appropriate. We thought
about measuring success through usage statistics (downloads, altmetrics).
``Should funders demand reproducibility?'' we wondered. In order for results
to be reproduced, software would need to be carefully curated.

>>>>

The Second Workshop on Sustainable Software for Science: Practice and
Experiences
(WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was
held on  16 November, 2014 in conjunction with the International
Conference for High Performance Computing, Networking, Storage and Analysis
(SC14)\footnote{\url{http://sc14.supercomputing.org}}. WSSSPE2 followed the
model of a general initial workshop,
WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
which co-occurred with SC13, and a focused workshop,
WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}},
which was organized in July 2014 jointly with the SciPy 
conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}}.

@@

The review process was fairly standard. First, reviewers bid for papers. Then an
automated system matched the bids to determine assignments. Next, reviewers 
completed their assigned reviews (with an average of 4.9 reviews per paper, and
4.1 reviews per reviewer). This process was done through
EasyChair\footnote{http://easychair.org/}, which allowed reviewers to provide
scores on relevance and comments to the organizers, who used the information to
decide which papers to associate with the workshop and provided the comments to
the authors to help them improve their papers.

@@

\subsection{Marlon Pierce: Patching It Up, Pulling It Forward~\cite{Pierce_poster}}
Pierce discussed how open open source is. Open software needs a diverse, openly
governed community behind it, just as it needs open licensing and a public code
repository. To probe the level of governance within open source projects, he and
his co-authors (Suresh Marru and Chris Mattmann) suggested a contest to encourage
individual developers to submit patches and pull request to projects that are
important to them. This simple mechanism will expose several governance
mechanisms, such as how easy it is for independent developers to communicate
with project leadership, how projects accept and properly license third-party
contributions, and how projects make decisions such as granting source tree
write access.

@@

Of the short actionable papers that would lead to improvements for sustainable
software science, 11 submissions were categorized in the  Software Development Experiences
group. Because of the large number, we split these into two subgroups
prior to the event. Some common themes helped in this division. For
example, several papers that addressed education and training issues,
including best practices and case studies, were grouped
together. Others discussed experiences with registries, developer
collectives and specific examples of successful, sustainable software
(including a valuable industry perspective).

@@

We asked ourselves, ``Where are there examples of things that are changing
because of this and how can we build momentum?'' One example demonstrates
change over time. In 2007, nanoHUB listed the academic reward structure as
a problem in an EDUCAUSE report where the authors note, ``In the future,
nanoHUB researchers are hoping to change the research culture. While they
recognize that young faculty members are unlikely to get tenure based on
their nanoHUB contributions, they hope to encourage faculty to think
beyond their own research needs to consider publishing tutorials and other
Web content in their field on the nanoHUB site.'' Fast forward seven years
later where, in a 2014 iSGTW article, quote nanoHUB Principal Investigator (PI) Gerhard Klimeck, ``A
former student of mine published eight tools on nanoHUB, serving over
6,000 people with his tools. He then joined a university as a professor
and introduced nanoHUB. Use of the gateway from that university
skyrocketed; he used nanoHUB in existing classes, created new classes, and
infused it in his research. Ultimately, the professor's department head
associates his two-year rise to tenure with the notoriety and innovation
he gained through nanoHUB.''

@@

We then moved beyond our training discussion to address funding that
encourages sustainable software -- funding of both people and projects
that create a true system to support sustainability. We called this
``institutionalized serendipity.'' As science is increasingly reliant
on software, one participant observed that ``software development can have
much broader impact than publishing individual research, but it is not
viewed that way.'' Because of this centrality, one participant mentioned
that training ought to be called Science Carpentry rather than Software
Carpentry. We felt we were beginning to see changes in the research
community as a result of the NSF's data management plan requirements. PIs 
are thinking more about data and some university libraries are offering
data repositories. We wondered if a software management plan might be
effective. ``How might funding programs need to adapt to reward good
software development practices?'' we asked ourselves. We thought about
best practices such as version control, test harnesses, mailing lists, bug
tracking, community contribution, and reuse where appropriate. We thought
about measuring success through usage statistics (downloads, altmetrics).
``Should funders demand reproducibility?'' we wondered. In order for results
to be reproduced, software would need to be carefully curated.
